---
title: "process lung autopsy all"
output: 
  html_document:
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE)
library(Seurat)
library(dplyr)
library(stringr)
library(ggplot2)
library(cowplot)
library(igraph)
library(Matrix)
library(tidyr)
options(future.globals.maxSize = 10 * 1024 * 1024^2)
```

results on lung autopsies with 10X platform. 8 samples (6 autopsy, 2 controls).  alignment + quantification using `cellranger` against GRCh38+viral genome. we take all cells with at least 500 genes and mark doublets called by [scrublet](https://github.com/swolock/scrublet).

```{r get_data,fig.width=8,fig.height=4}
samples <- c('A333-chronic',
             'A344-R3-chronic',
             'A592-prolonged',
             'BB73-control',
             'O325-akut',
             '17B-control',
             'T128-akut',
             'T188-R2-prolonged')

source('ReadCB_h5.R')

sobjs <- list()
for (sample in samples) {
  h5_file <- file.path('data','autopsy','cellbender',paste0(sample,'_filtered.h5'))
  if (file.exists(h5_file)) {
    dge <- ReadCB_h5(h5_file)
    colnames(dge) <- paste0(gsub('_','-',sample),'_',colnames(dge))
    scrublet <- read.csv(file.path('data','autopsy','scrublet',paste0(sample,'_scrublet.csv')),row.names=1)
    row.names(scrublet) <- paste0(gsub('_','-',sample),'_',row.names(scrublet))
    sobjs[[sample]] <- CreateSeuratObject(dge,project=sample,
                                          assay='RNA',min.cells=1,min.features=500, 
                                          meta.data=scrublet)
  }
}

sobj <- merge(sobjs[[1]],sobjs[2:length(samples)])

sobj[['donor']] <- gsub('\\-.*','',sobj$orig.ident)
sobj[['condition']] <- gsub('.*\\-([^-]*$)','\\1',sobj$orig.ident)
sobj[['origin']] <- 'autopsy'

sobj[['pct.mito']] <- PercentageFeatureSet(sobj, pattern='^MT-')
sobj[['pct.ribo']] <- PercentageFeatureSet(sobj, pattern='^RP[SL][0-9]*$')
sobj[['pct.MALAT1']] <- PercentageFeatureSet(sobj, pattern='^MALAT1$')
sobj[['pct.SCoV2']] <- PercentageFeatureSet(sobj, pattern='^SCoV2-')
```

here's some basic statistics. 

```{r stats,fig.width=8,fig.height=2}
sobj@meta.data %>%
  dplyr::group_by(orig.ident) %>%
  dplyr::summarise(ncells=n(),
                   ndoublets=sum(scrublet_prediction=='doublet'),
                   nGene=median(nFeature_RNA),
                   nUMI=median(nCount_RNA),
                   pct.mito=mean(pct.mito),
                   pct.ribo=mean(pct.ribo),
                   pct.MALAT1=mean(pct.MALAT1),
                   pct.SCoV2=mean(pct.SCoV2))

ggplot(sobj@meta.data %>% 
         dplyr::select(orig.ident, nCount_RNA, nFeature_RNA,  pct.mito, 
                       pct.ribo, pct.MALAT1, pct.SCoV2) %>%
         gather(metric,value,-orig.ident),
       aes(x=orig.ident,y=value,fill=orig.ident)) + 
  geom_boxplot(outlier.size=.5) + 
  facet_wrap(~metric,ncol=6,scales='free_y') + 
  theme(axis.text.x=element_blank())
```

perform standard Seurat workflow, keeping cells with less than 10% mitochondrial genes and less than 5000 genes and remove doublets

```{r process}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

sobj <- subset(sobj, (pct.mito < 10) & (nFeature_RNA < 5000) & (scrublet_prediction!='doublet')) %>%
  NormalizeData(normalization.method='LogNormalize',scale.factor=10000,verbose=FALSE) %>%
  FindVariableFeatures(selection.method='vst',nfeatures=2000,verbose=FALSE) %>%
  ScaleData(vars.to.regress=c('nCount_RNA'),verbose=FALSE) %>%
  CellCycleScoring(s.features = s.genes, g2m.features = g2m.genes) %>%
  RunPCA(features=VariableFeatures(.),verbose=FALSE) %>%
  FindNeighbors(dims=1:20,verbose=FALSE) %>%
  RunUMAP(dims=1:20,verbose=FALSE)
```

there's some variability between the two donors.

```{r show,fig.width=6,fig.height=3.5}
DimPlot(sobj,reduction='umap',group.by='donor')
```

we remove the batch effect with `IntegrateData`

```{r integrate}
sobjs <- SplitObject(sobj, split.by='donor')
anchors <- FindIntegrationAnchors(sobjs, dims=1:20, verbose=FALSE)
sobj <- IntegrateData(anchors, dims=1:20, verbose=FALSE)
DefaultAssay(sobj) <- "integrated"
sobj <- ScaleData(sobj,
                  verbose = FALSE) %>%
  RunPCA(npcs = 20, verbose = FALSE) %>%
  FindNeighbors(dims=1:20, verbose = FALSE) %>%
  RunUMAP(dims = 1:20,verbose = FALSE) %>%
  FindClusters(verbose=FALSE)
```

the same plots as before, now better overlap between donors

```{r show_integrated,fig.width=6,fig.height=3.5}
DimPlot(sobj,reduction='umap',group.by='donor')
DimPlot(sobj,reduction='umap',group.by='condition')
DimPlot(sobj,reduction='umap',group.by='Phase')
DimPlot(sobj,reduction='umap',label=TRUE)
```

we use the dataset by Travaglini et al. (bioRxiv https://doi.org/10.1101/742320 ) to score cell types. 

```{r score_celltypes,fig.height=12,fig.width=10}
travaglini <- readRDS('data/seurat/Travaglini.rds')
DefaultAssay(sobj) <- 'integrated'
anchors <- FindTransferAnchors(reference = travaglini, query = sobj, dims = 1:20,
                               features = intersect(row.names(travaglini@assays$RNA@scale.data),
                                                    row.names(sobj@assays$integrated@scale.data)))
pred <- TransferData(anchorset = anchors, refdata = travaglini$free_annotation, dims = 1:20)
sobj <- AddMetaData(sobj, metadata=pred)
sobj@meta.data %>%
  dplyr::group_by(seurat_clusters,predicted.id) %>%
  dplyr::summarise(n=n()) %>%
  dplyr::mutate(frac=round(n/sum(n),2)) %>%
  dplyr::filter(frac > .1)
CombinePlots(plots=list(DimPlot(sobj, group.by='predicted.id', label=TRUE, repel=TRUE),
                        FeaturePlot(sobj,'prediction.score.max', label=TRUE),
                        DimPlot(sobj, label=TRUE)),ncol=1)
```

use DoubletFinder to estimate doublets

```{r estimate_doublets,fig.width=7,fig.height=4}
source('my_doubletFinder.R')
DF <- list()
for (sample in samples) {
  tmp <- subset(sobj,subset=orig.ident==sample)
  DefaultAssay(tmp) <- 'RNA'
  homotypic.prop <- modelHomotypic(Idents(tmp))
  nExp_poi <- round(0.05*nrow(tmp@meta.data)) 
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  tmp <- my_doubletFinder(tmp, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi, reuse.pANN = FALSE)
  tmp <- my_doubletFinder(tmp, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi.adj, reuse.pANN = "pANN")
  DF[[sample]] <- tmp@meta.data[,'DF.classifications',drop=FALSE] %>%
    tibble::rownames_to_column('cell')
}
sobj <- AddMetaData(sobj, do.call(rbind,DF) %>% tibble::remove_rownames() %>% tibble::column_to_rownames('cell'))
DimPlot(sobj,group.by='DF.classifications')
table(sobj$DF.classifications)
```

```{r save_data}
saveRDS(sobj,file='data/seurat/lung_autopsy_all.rds')
```

```{r sessionInfo}
sessionInfo()
```